---
title: "A different analysis of translation efficiency"
author: ""
date: ""
output:
    html_document:
        toc: true
        theme: united
    pdf_document:
        toc: true
        latex_engine: xelatex
---

```{r load-libraries, echo=FALSE, message=FALSE}
modules::import('scripts/knit')
tidyr = import_package('tidyr')
```

# The problem

```{r load-old-te-data, echo=FALSE}
te = readRDS('./results/te-human.rds') %>% filter(Type %in% c('Matching', 'DE'))
n_samples = nrow(filter(te, Type == 'Matching'))
```

Our current analysis for Figure 3 fundamentally suffers from a lack of
statistical power.

```{r problematic-figure, echo=FALSE, fig.width=3, fig.height=3, fig.align='center'}
ggplot(te) +
    aes(x = Type, y = TE) +
    geom_boxplot() +
    geom_point(position = position_jitter(0.2))
```

---

# The problem

* The number of samples (`r n_samples`) is low
* We are using a nonparametric test of different medians
* The distributions are not continuous
* There are ties, so we cannot compute exact $p$-values:

    \footnotesize
    ```{r}
    wilcox.test(TE ~ Type, te, alternative = 'less')
    ```

→ We lack statistical power to establish lack of evidence convincingly. This
can only be solved by increasing the number of sequenced libraries
dramatically.

---

# The problem

* *Finally*, we’re arguably testing the wrong thing:

    * We are testing whether cell-type specific genes are significantly more
      efficiently translated than the bulk of the transcriptome.

    * But we *want* to test whether cell-type specific genes are significantly
      more efficiently translated by the cell-type specific tRNA pool, compared
      to mismatching tRNA pools.

* With this, I propose two changes:

> **Note** All data in the subsequent plots is mock data!

---

# Compare matching tRNA pool to mismatching tRNA pool.

For each test set, compare the TE calculated from matching and mismatching tRNA
pools:

```{r example-data, echo=FALSE}
set.seed(24587954)
conditions = factor(c('All', 'DE', 'GO', 'Ribosomal'), ordered = TRUE)
make_test_set = size ~ cond -> data_frame(TE = rnorm(size, 0.7, 0.2), Type = cond)
matching = bind_rows(lapply(conditions, cond -> make_test_set(14, cond)))
mismatching = bind_rows(lapply(conditions, cond -> make_test_set(20, cond)))
new_te = bind_rows(mutate(matching, `tRNA pool` = 'Match'),
                   mutate(mismatching, `tRNA pool` = 'Mismatch'))
```

```{r boxplot-te, echo=FALSE, fig.width=5, fig.height=3, fig.align = 'center'}
ggplot(new_te, aes(x = `tRNA pool`, y = TE)) +
    geom_boxplot(aes(color = `tRNA pool`)) +
    scale_x_discrete('', labels = '') +
    facet_wrap(~ Type, ncol = 4)
```

---

# Compare matching tRNA pool to mismatching tRNA pool.

* This test is more direct. However, it does not fundamentally solve the fact
  that we are trying to *accept* the null hypothesis of no difference, rather
  than rejecting it. All we can therefore say is that we find no evidence
  (given insufficient data).

* An even more direct test is needed, and possible:

---

# Correlate TE of matching and mismatching tRNA pool

Here’s a different idea:

* For each gene set, say $\text{DE} = (g_1, g_2, \dots, g_n)$, calculate
  per-gene TE with **matching** tRNA pool

    $$t = (t_{g_1}, t_{g_2}, \dots, t_{g_n})$$

    and **mismatching** tRNA pool

    $$t' = (t_{g_1}, t_{g_2}, \dots, t_{g_n})$$

* Calculate the correlation
  
    $$\rho = cor(t, t')$$

---

# Correlate TE of matching and mismatching tRNA pool

```{r example-de-data, echo=FALSE}
t_de_match = rnorm(200, 0.7, 0.2)
t_de_mismatch = t_de_match + rnorm(200, 0.1, 0.2)
t_de = bind_rows(data_frame(Gene = 1 : 200, TE = t_de_match, tRNA = 'Match'),
                 data_frame(Gene = 1 : 200, TE = t_de_mismatch, tRNA = 'Mismatch'))
```

```{r te_de_scatter, echo=FALSE, fig.width=3, fig.height=3, fig.align='center'}
t_de_untidy = tidyr$spread(t_de, tRNA, TE)
t_rho = cor(t_de_untidy$Match, t_de_untidy$Mismatch, method = 'spearman')
ggplot(t_de_untidy, aes(x = Match, y = Mismatch)) +
    geom_point() +
    geom_smooth(method = lm) +
    annotate(geom = 'text', x = 1, y = 0,
             label = paste('rho[DE] ==', format(t_rho, digits = 2)), parse = TRUE)
```

---

# Correlate TE of matching and mismatching tRNA pool

* If there’s *no* adaptation of TE to cell-specific tRNA pools, then $\rho$ is
  higher than expected by chance.

* To quantify what’s expected by chance,

    * simulate tRNA pools, calculate TE using these pools as $t_{1\dots n}^*$,

    * calculate $\rho_{1 \dots n}^*$, for each simulated pool, and

    * calculate empirical one-tailed $p$-value of $\rho \sim \rho^*$.

* The above was simplified, we have no 1:1 match between “match” and
  “mismatch” tRNA:

    * We have 3 matching TEs, “liver”, “HepG2” and “Huh7”
    * We have 4 mismatching TEs: “liver–HepG2”, “liver–Huh7”,
    “HepG2–liver”, “Huh7–liver”

    **How to handle these?**

---

# Details

Empirically testing $\rho \sim \rho^*$ is impossible for a single $\rho$ but
since we have more than one match- and mismatch correlation we effectively
have sample of $\rho$ values; we can consequently use a Kolmogorov–Smirnov
test to determine if the sample of $\rho$ was drawn from $\rho^*$.
