---
title: "GC bias and CUB conservation"
author: "klmr"
date: "15 January 2015"
output: beamer_presentation
---

```{r echo=FALSE}
load('slides.rdata')
```

## Summary

We observe that GC bias is highly correlated with PC1 of the codon usage per GO
term PCA.

```{r, echo=FALSE}
plot(pc_cu_by_aa$x[, 1], go_gc, xlab = 'PC1 of per-amino acid codon usage bias',
     ylab = 'GC bias',
     main = 'Correlation of codon usage bias and GC bias per GO term')
```

---

We want to compare whether GC bias is, in fact, subject to stronger
evolutionary conservation than codon usage across GO terms.

In other words, compare conservation (across species) of codon usage bias and GC
bias for each GO term; and determine whether GC bias conservation is
significantly different:

---

### Species $n$ ($S_n$)

 &nbsp;   | Gene 1   | Gene 2   | $\dots$  | Gene $k$
--------- | -------- | -------- | -------- | --------
Codon 1   | $c_{11}$ | $c_{12}$ | $\dots$  | $c_{1k}$
Codon 2   | $c_{21}$ | $c_{22}$ | $\dots$  | $c_{1k}$
$\vdots$  | $\vdots$ | $\vdots$ | $\ddots$ | $\vdots$
Codon $m$ | $c_{m1}$ | $c_{m2}$ | $\dots$  | $c_{mk}$
GC bias   | $gc_1$   | $gc_2$   | $\dots$  | $gc_k$

---

### Correlations

We can now calculate correlations

$$\{Cor_{c_{1.}}(S_1, S_2), Cor_{c_{2.}}(S_1, S_2), \dots, Cor_{GC}(S_1, S_2)\}$$

That way, we can calculate an empirical $p$-value for the null hypothesis

> **$H_0$**: the correlation across species of the GC bias is not higher than
> the correlation across species of individual codons.

We can do this for each GO term, and test for an enrichment of significant $p$s.

## Potential issues

* For each gene, the GC bias is correlated with the codon distribution
* For each gene, codon usage biases are correlated amongst each other

As a consequence, we don’t know whether changing codon usage between species has
implications for the conservation of the GC bias.

---

To solve this, we can simulate different species by mutating the codon usage of
genes

1. freely (but neutrally, i.e. preserving amino acid identity)
2. under the constraint that the GC bias remains the same

And assess how this impacts the cross-species correlation of codon usage and GC
bias.

Finally, we will be able to see whether GC bias conservation between species is
systematically larger than between simulated ones.
