```{r echo=FALSE}
options(stringsAsFactors = FALSE)
library(dplyr)
library(magrittr)
```
Read pol III tRNA counts.

```{r}
cancer_counts = read.csv('./data/pol3-chip-trna-counts-cancer.tsv', sep = '\t', row.names = 1)

head(cancer_counts)
```

Read sample library metadata.

```{r}
cancer_libraries = read.csv('./data/pol3-chip-cancer-libraries.tsv') %>%
    select(DO = 1, Type = 2, Celltype = 3) %>%
    filter(DO %in% colnames(cancer_counts))

cancer_libraries
```

Show sample distributions.

```{r}
# Reorder columns to match `cancer_libraries`
cancer_counts = cancer_counts[, cancer_libraries$DO]

eps = min(cancer_counts[cancer_counts > 0])
boxplot(cancer_counts + eps, log = 'y',
        col = ifelse(cancer_libraries$Type == 'input', 'gray', 'white'))
```

Normalise by library size.

```{r}
library(DESeq2)

cancer_libraries %<>% filter(Type == 'ChIP') %>%
    select(-Type)

cancer_counts_chip = select(cancer_counts, one_of(cancer_libraries$DO))
cancer_size_factors = estimateSizeFactorsForMatrix(cancer_counts_chip)
cancer_counts_chip = t(t(cancer_counts_chip) / cancer_size_factors) %>%
    as.data.frame
```

Plot normalised counts.

```{r}
eps = min(cancer_counts_chip[cancer_counts_chip > 0])
boxplot(cancer_counts_chip + eps, log = 'y')
```

Replicate variability: log-log-plot of counts.

```{r}
library(ggplot2)

plot_compare_replicates = function (data, cols, main) {
    do_plot = function (data, cols, main) {
        if (missing(main))
            main = sprintf('Replicate variability between %s and %s',
                           cols[1], cols[2])
        p = ggplot(data, aes_string(x = cols[1], y = cols[2])) +
            geom_point() +
            scale_x_log10() +
            scale_y_log10() +
            ggtitle(main)
        plot(p)
    }

    cols = if (is.list(cols)) cols else list(cols)

    if (missing(main))
        mapply(do_plot, list(data), cols)
    else
        mapply(do_plot, list(data), cols, main)

    invisible()
}

cancer_libraries %>%
    group_by(Celltype) %>% do(Replicates = .$DO) %T>%
    {plot_compare_replicates(cancer_counts_chip, .$Replicates,
                             unique(.$Celltype))} %>%
    invisible
```

Load the mouse development data.

```{r}
devo_libraries = read.csv('./data/trna/chip-sample-map.tsv', sep = '\t',
                          header = FALSE) %>%
    select(DO = 1, Type = 2, Tissue = 3, Stage = 4) %>%
    mutate(DO = paste0('do', DO), Stage = toupper(Stage)) %>%
    filter(Type == 'PolIII', Tissue == 'liver') %>%
    select(-Type, -Tissue)

devo_counts = read.csv('./data/trna/pol3-trna-counts.tsv', sep = '\t')
devo_counts = devo_counts[, devo_libraries$DO]
head(devo_counts)
```

Build DESeq experiment, comparing cancers against E15.5.

```{r}
devo_libraries %<>% filter(Stage == 'E15.5') %>%
    dplyr::rename(Condition = Stage)
cancer_libraries %<>% dplyr::rename(Condition = Celltype)
experimental_design = rbind(devo_libraries, cancer_libraries)
raw_counts_table = cbind(devo_counts, cancer_counts)[, experimental_design$DO]
size_factors = estimateSizeFactorsForMatrix(raw_counts_table)
normalized_counts = t(t(raw_counts_table) / size_factors)
```

Compare the samples

```{r fig.width=9}
cor = cor(normalized_counts, method = 'spearman')

library(gplots)
divergent_colors = colorRampPalette(c('#603D71', 'white', '#A4B962'))(30)
celltype_colors = c(E15.5 = 'chartreuse4',
                    `Hepa1-6` = 'dodgerblue3',
                    Hepa1c1c7 = 'deepskyblue3')
heatmap.2(cor, trace = 'none', density.info = 'none', col = divergent_colors,
          ColSideColors = celltype_colors[experimental_design$Condition],
          lmat = cbind(c(5, 5, 3), c(4, 1, 2), c(6, 6, 6)),
          lwid = c(1.5, 4, 2),
          lhei = c(1.5, 0.25, 4))
legend('right', bty = 'n', legend = names(celltype_colors),
       fill = celltype_colors)
```

Perform differential expression analysis between all pairwise conditions: in
particular, we want to know what changes between cancer and developmental
tissue; however, the between-cancer comparison may yield insight into the
inherent variability of the different conditions.

```{r}
de = function (data, col_data, contrast) {
    col_data %<>% filter(Condition %in% contrast)
    cds = DESeqDataSetFromMatrix(data[, col_data$DO], col_data, ~Condition)
    cds$Condition = relevel(cds$Condition, contrast[1])
    DESeq(cds)
}

summarize_de = function (cds) {
    contrast = levels(cds$Condition)
    main = function (title)
        sprintf('%s between %s and %s',
                title, contrast[1], contrast[2])
    plotDispEsts(cds, main = main('Dispersion estimates'))
    plotMA(results(cds), alpha = 0.01, main = main('Differential expression'))
}

cds_e155_hepa16 = de(raw_counts_table, experimental_design,
                     c('E15.5', 'Hepa1-6'))

summarize_de(cds_e155_hepa16)

cds_e155_hepa1c1c7 = de(raw_counts_table, experimental_design,
                        c('E15.5', 'Hepa1c1c7'))

summarize_de(cds_e155_hepa1c1c7)

cds_hepa = de(raw_counts_table, experimental_design,
              c('Hepa1-6', 'Hepa1c1c7'))

summarize_de(cds_hepa)
```
