```{r echo=FALSE}
modules::import('./scripts/knit', attach = TRUE)
opts_chunk$set(fig.path = 'figure/sample-size-effect-',
               cache.path = 'cache/sample-size-effect-',
               dev = c('png', 'pdf'))

config = modules::import('./config_human')
```

# The effect of sample size on properties of genes

Here we want to test how a certain feature of gene sets (in fact, the mean
codon usage, CU) is affected by random sampling. To test this, we sample
random sets of genes of different sizes and observe the effect this has on the
codon usage.

To make these results comparable with relevant data, we draw samples with
sizes equal to those of all GO categories (over some minimal size threshold).
Consequently, the first step is to load the GO terms.

We ignore very small GO term sizes (less than 40) as well as very large GO term
sizes (greater than 10000).

```{r go_terms}
go_terms = io$read_table('./data/gene_association.goa_human', sep = '\t',
                         quote = '', comment.char = '!', header = FALSE) %>%
    select(GeneName = 3, GO = 5) %>%
    tbl_df()

mrna_annotation = io$read_table(config$mrna_annotation, header = TRUE) %>%
    filter(source == 'protein_coding') %>%
    select(Gene = ID, Name) %>%
    tbl_df()

go_terms = inner_join(go_terms, mrna_annotation, by = c('GeneName' = 'Name'))

distinct_go_sizes = go_terms %>% group_by(GO) %>%
    summarize(Size = n()) %>%
    filter(Size >= 40 & Size <= 10000) %>%
    .$Size %>%
    unique() %>%
    sort()
```

For these sizes, we sample random genes to calculate the CU for. This requires
us to load the coding sequences of those genes.

```{r cds}
bios = loadNamespace('Biostrings')
cds = bios$readDNAStringSet(config$cds)
names(cds) = sub('.*gene:(ENS(MUS)?G\\d+).*', '\\1', names(cds))
cds = data.frame(Gene = names(cds), Sequence = as.character(cds))

# Filter CCDS, only preserve valid coding frames

is_valid_cds = function (seq)
    nchar(seq) %% 3 == 0 & grepl('^ATG', seq) & grepl('(TAG|TAA|TGA)$', seq)

cds = cds %>% filter(is_valid_cds(Sequence))
canonical_cds = cds %>%
    group_by(Gene) %>%
    arrange(desc(nchar(Sequence))) %>%
    dplyr::slice(1)
```

The next steps require the genetic code, without stop codons.

```{r genetic_code}
full_genetic_code = cbind(AA = bios$GENETIC_CODE,
                          Codon = names(bios$GENETIC_CODE)) %>% as.data.frame()
genetic_code = filter(full_genetic_code, AA != '*')
```

Now we calculate the individual codon usage for each gene.

```{r codon_usage}
codon_usage = canonical_cds$Sequence %>%
    bios$DNAStringSet() %>%
    bios$trinucleotideFrequency(3) %>%
    as.data.frame() %>%
    {cbind(Gene = canonical_cds$Gene, .)} %>%
    melt(id.vars = 'Gene', variable.name = 'Codon', value.name = 'Count') %>%
    inner_join(genetic_code, by = 'Codon') %>%
    tbl_df()
```

… and then the genomic background CU, which we’ll subsequently use as a point
of comparison, to assess variability of CU for different sample sizes.

**Caveat** In the following, we will use the codon usage, normalised by the
overall codon count (such that the values for all codons sum to 1). This is
done to make CU between different gene sets of different size comparable.
Initially I calculated the *relative* codon usage (RCU), which normalises each
codon frequency by the sum of the codon frequency of its synonymous codons. This
removes the amino acid usage bias as a confounder. However, it turns out that
this makes all RCUs of different gene sets highly correlated, since, across all
amino acids, codons with many synonymous codons will have lower values compared
to codons with few synonymous codons (simply because usage of synonymous codons
sums to 1).

```{r background_cu}
background_cu = codon_usage %>%
    group_by(AA, Codon) %>%
    summarize(Count = sum(Count)) %>%
    ungroup() %>%
    mutate(Prop = Count / sum(Count)) %>%
    arrange(Codon)
```

The `arrange` statement above serves to make the order of the CU guaranteed so
that we can directly correlate it with the sampled CU, calculated next. This
is probably unnecessary (and time-consuming) but I cannot find anything in the
dplyr documentation guaranteeing the order, and better safe than sorry.

Now we sample repeatedly for each set size, and calculate the mean relative
codon usage (mean CU), which we then correlate with the background CU to
assess codon usage variability.

```{r cu_fit_function}
# Optimised to allow indexed access for gene names, since selection by gene name
# turned out to be a major bottleneck.
# This change provides a > 1000% speedup.
stride = length(unique(codon_usage$Gene))
stopifnot(stride * 61 == nrow(codon_usage))
stopifnot(all(codon_usage[1 : stride, ]$Codon == codon_usage$Codon[1]))

cu_fit = function (gene_index_set) {
    translated_index = unlist(lapply(gene_index_set, i -> 0 : 60 * stride + i))
    codon_usage[translated_index, ] %>%
        group_by(AA, Codon) %>%
        summarize(Count = sum(Count)) %>%
        ungroup() %>%
        mutate(Prop = Count / sum(Count)) %>%
        arrange(Codon) %>%
        {cor(.$Prop, background_cu$Prop)}
}
```

```{r sampled_cu_fit, eval=FALSE}
rng_seed = 1428079834

sample_cu_fit = function (size)
    cu_fit(sample.int(nrow(canonical_cds), size))

sample_cu_fit_rep = function (size, repetitions = 2) {
    # Ensure that all simulations are using different seed, as otherwise the
    # parallel jobs will start off with the same sequence of random samples.
    set.seed(rng_seed + size)
    on.exit(cat('.', file = stderr()))
    replicate(repetitions, sample_cu_fit(size))
}

cores = parallel::detectCores()
sampled_cu_fit = parallel::mclapply(distinct_go_sizes, sample_cu_fit_rep,
                                    mc.cores = cores, mc.set.seed = FALSE) %>%
    setNames(distinct_go_sizes) %>%
    {do.call(rbind, .)} %>%
    as.data.frame() %>%
    add_rownames('Size')

cat('\n', file = stderr())
```

```{r save_sampled_data, echo=FALSE, eval=FALSE}
saveRDS(sampled_cu_fit, file = './results/sampled-cu-fit.rds')
```

```{r load_sampled_data, echo=FALSE}
# The above calculation is *hellishly* slow – it took 230 CPU hours to complete.
# Consequently, we ran it on a cluster and load the results here, instead of
# actually computing them.
sampled_cu_fit = readRDS('./results/sampled-cu-fit.rds')
```

Bring the data into a more convenient format.

```{r melt_sampled_data}
sampled_cu_fit = sampled_cu_fit %>%
    melt(id.vars = 'Size', value.name = 'Correlation') %>%
    mutate(Size = as.numeric(Size)) %>%
    select(-2) %>%
    tbl_df()
```

Plot histograms for the distribution. Take out highest observations because they
are useless and skew everything.

```{r sampled_data_plot}
sampled_summary = sampled_cu_fit %>%
    group_by(Size) %>%
    do(.$Correlation %>% quantile(c(0, 0.02, 0.25, 0.5, 0.75, 0.98, 1)) %>%
       unclass() %>% as.data.frame() %>% add_rownames('Point')) %>%
    dcast(Size ~ Point, value.var = '.') %>%
    tbl_df() %>%
    filter(Size < 10000)

# We extent whiskers outside by 1.5 IQR from the IQR. This is a simplification
# which is aceptable here since the data is expected to be dense around that
# point so searching for an infimum/supremum brings little added accuracy.
ggplot(sampled_summary) +
    geom_segment(aes(x = Size, y = `25%`,
                     xend = Size, yend = `75%`),
                 color = '#808080') +
    geom_point(aes(x = Size, y = `2%`), size = 0.5)  +
    geom_point(aes(x = Size, y = `50%`), size = 1) +
    geom_point(aes(x = Size, y = `98%`), size = 0.5)  +
    scale_x_log10(name = 'Gene set size',
                  breaks = c(50, 100, 500, 1000, 5000, 10000)) +
    scale_y_log10(name = 'Codon usage correlation',
                  breaks = seq(0.89, 1, by = 0.01), limits = c(0.89, 1))
```

Now calculate the GO term CU and its correlation with the genomic CU.

```{r go_cu_fit}
# Index gene name => `codon_usage` row index, necessary for lookup.
gene_index = data.frame(Gene = codon_usage[seq_len(stride), ]$Gene,
                        Index = seq_len(stride))

# We are only interested in bigger GO term gene sets.
big_go_terms = go_terms %>%
    group_by(GO) %>%
    filter(n() >= 40) %>%
    inner_join(gene_index, by = 'Gene')

# Since a join over > 1M rows becomes prohibitively slow, we use a loop +
# `filter` here instead.
go_term_ids = unique(big_go_terms$GO)

go_cu_fit = function (go_id)
    cu_fit(filter(big_go_terms, GO == go_id)$Index)

go_cu = parallel::mclapply(go_term_ids, go_cu_fit, mc.cores = cores) %>%
    setNames(go_term_ids) %>%
    unlist()

go_sizes = big_go_terms %>%
    group_by(GO) %>%
    summarize(Size = n())

go_cu_by_size = data.frame(GO = names(go_cu), Correlation = go_cu) %>%
    inner_join(go_sizes, by = 'GO') %>%
    tbl_df()
```

```{r go_cu_fit_plot}
#ggplot(go_cu_by_size, aes(x = factor(GO, GO[order(Size)], ordered = TRUE),
ggplot(go_cu_by_size, aes(x = Size, y = Correlation)) +
    geom_point() +
    scale_x_log10()
```
