# Measures of translation efficiency

The following uses expression data from human liver. We will look at codon
usage of GO terms.

```{r load-data}
modules::import('./scripts/knit', attach = TRUE)
config = import('./config_human')
data = import('./data')

go_genes = data$go_genes(config)
canonical_cds = data$canonical_cds(config)

mrna_library = data$mrna_design(config) %>%
    filter(Celltype == 'Liver-Adult') %>% .$DO %>% first()
mrna_counts = data$mrna_counts(config) %>%
    select_(~Gene, ~Name, Count = mrna_library) %>%
    inner_join(canonical_cds, by = 'Gene')

trna_library = data$trna_design(config) %>%
    filter(Celltype == 'Liver-Adult') %>% .$DO %>% first()
trna_counts = data$trna_counts(config) %>%
    select_(~Gene, AminoAcid = ~AA, ~Anticodon, Count = trna_library)
```

“Translation efficiency” (TE) measures how well codon usage and anticodon
abundance are adapted. Specifically, whether a given codon can be efficiently
translated because its concomitant tRNA anticodon is efficiently found by the
ribosome. This requires the tRNA to be abundant in just the right ratio — that
dictated by codon usage, relative to the usage of other codons.

We therefore scale codon usage (CU) and anticodon abundance (AA) by the total
codon usage and anticodon abundance.

```{r cu}
cu_ = import('codon_usage')
cu = mrna_counts %>%
    cu_$cu() %>%
    inner_join(go_genes, by = 'Gene') %>%
    group_by(GO, Codon) %>%
    summarize(CU = sum(CU)) %>%
    mutate(CU = CU / sum(CU))

aa = trna_counts %>%
    group_by(Anticodon) %>%
    summarize(AA = sum(Count)) %>%
    mutate(AA = AA / sum(AA)) %>%
    mutate(Codon = cu_$revcomp(Anticodon))

te_cu = inner_join(cu, aa, by = 'Codon') %>%
    summarize(TE = cor(CU, AA, method = 'spearman'))
```

A different approach is to look at the selective pressure exerted on the codon
deployment: that is, how abundant a given codon is compared to synonymous codons
(ignoring wobble base pairing for now). This compels us to look at relative
codon usage (RCU) within a synonymous codon family and, conversely, at the
relative anticodon abundance (RAA) within a tRNA anticodon isoacceptor family.

```{r rcu}
rcu = mrna_counts %>%
    cu_$cu() %>%
    inner_join(cu_$genetic_code, by = 'Codon') %>%
    inner_join(go_genes, by = 'Gene') %>%
    group_by(GO, AA, Codon) %>%
    summarize(CU = sum(CU)) %>%
    mutate(RCU = CU / sum(CU)) %>%
    group_by(GO)

raa = trna_counts %>%
    group_by(AminoAcid, Anticodon) %>%
    summarize(RAA = sum(Count)) %>%
    mutate(RAA = RAA / sum(RAA)) %>%
    ungroup() %>%
    mutate(Codon = cu_$revcomp(Anticodon))

te_rcu = inner_join(rcu, raa, by = 'Codon') %>%
    summarize(TE = cor(RCU, RAA, method = 'spearman'))
```

Plot the distributions

```{r plot-te}
te = bind_rows(mutate(te_cu, Mode = 'CU vs AA'),
               mutate(te_rcu, Mode = 'RCU vs RAA'))

ggplot(te) + geom_boxplot(aes(Mode, TE))
```
